---
title: "Airline_Traffic_TS"
output: html_document
date: "2024-03-27"
---

This is the Airline Traffic Time Series Project.

By: Harshith Samayamantula, Devan Pandya
3/27/24
Time Series 

```{r}
require(zoo)
require(ggplot2)
require(forecast)
require(TTR)
library(fpp2)
require(lubridate)
library(tseries)
library(urca)

airline_dataset = read.csv("air traffic.csv")
```

# Data Analysis Plan

## Research Question
Some research questions we want to explore include the following:
- Can we accurately forecast future air travel demand based on historical data? Specifically, passenger numbers and revenue passenger-miles?
  - The result is testing various methods for forecasting accuracy and discussing the level of accuracy we're able to attain.
- Is there a correlation between the number of flights (Flt) and total passenger volume (Pax)? How do revenue passenger-miles (RPM) correlate with the number of flights and passenger volume?
  - Correlation will be shown through autocorrelation and partial autocorrelation analysis, identifying if the data is stationary, and running other tests.

## Description

The air traffic dataset comes from Kaggle. It contains non-seasonally adjusted monthly air traffic data for all commercial U.S. air carriers from January 2003 to September 2023. It is being used under the US Government Works License and is collected by the US Department of Transportation Bureau of Transportation Statistics. The variables included in the dataset are listed as follows:

1. Year: Year
2. Month: Month
3. Dom_Pax: Domestic Air Travel Passengers
4. Int_Pax: International Air Travel Passengers
5. Pax: Total Air Travel Passengers
6. Dom_Flt: Number of Flights (Domestic)
7. Int_Flt: Number of Flights (International)
8. Flt: Number of Flights (Total)
9. Dom_RPM: Revenue Passenger Miles (Domestic)
10. Int_RPM: Revenue Passenger Miles (International)
11. RPM: Revenue Passenger Miles (Total)
12. Dom_ASM: Available Seat Miles (Domestic)
13. Int_ASM: Available Seat Miles (International)
14. ASM: Available Seat Miles (Total)
15. Dom_LF: Load Factor (Domestic)
16. Int_LF: Load Factor (International)
17. LF: Load Factor (Total)

Revenue Passenger-miles = the number of passengers and the distance flown in thousands (000)
Available Seat-miles = the number of seats and the distance flown in thousands (000)
Load Factor = passenger-miles as a proportion of available seat-miles in percent (%)

## Data Analysis 

#### Cleaning and Verification

```{r}
print(airline_dataset)

```

```{r}
colSums(is.na(airline_dataset))
```

```{r}
#Check for duplicated values
sum(duplicated(airline_dataset))

#Check for empty values
colSums(airline_dataset == "")
```

```{r}

airline_dataset$Year_Month = as.yearmon(paste(airline_dataset$Year, airline_dataset$Month), "%Y %m")

```

```{r}
print(airline_dataset)
str(airline_dataset)
```
```{r}
# Convert character columns to numeric columns
airline_dataset$Dom_Pax <- as.numeric(gsub(",", "", airline_dataset$Dom_Pax))
airline_dataset$Int_Pax <- as.numeric(gsub(",", "", airline_dataset$Int_Pax))
airline_dataset$Pax <- as.numeric(gsub(",", "", airline_dataset$Pax))
airline_dataset$Dom_Flt <- as.numeric(gsub(",", "", airline_dataset$Dom_Flt))
airline_dataset$Int_Flt <- as.numeric(gsub(",", "", airline_dataset$Int_Flt))
airline_dataset$Flt <- as.numeric(gsub(",", "", airline_dataset$Flt))
airline_dataset$Dom_RPM <- as.numeric(gsub(",", "", airline_dataset$Dom_RPM))
airline_dataset$Int_RPM <- as.numeric(gsub(",", "", airline_dataset$Int_RPM))
airline_dataset$RPM <- as.numeric(gsub(",", "", airline_dataset$RPM))
airline_dataset$Dom_ASM <- as.numeric(gsub(",", "", airline_dataset$Dom_ASM))
airline_dataset$Int_ASM <- as.numeric(gsub(",", "", airline_dataset$Int_ASM))
airline_dataset$ASM <- as.numeric(gsub(",", "", airline_dataset$ASM))
str(airline_dataset)
```
#### Converting to TS
```{r}
ts_airline <- ts(airline_dataset[, c("Dom_Pax", "Int_Pax", "Pax", "Dom_Flt", "Int_Flt", "Flt", "Dom_RPM", "Int_RPM", "RPM", "Dom_ASM", "Int_ASM", "ASM", "Dom_LF", "Int_LF", "LF")], start = c(2003, 1), frequency = 12)
```

```{r}
print(window(ts_airline, start = c(2003, 1), end = c(2003, 5)))
```
#### Visualizing Data

```{r}
par(mar = c(3, 3, 1, 1))
par(mfrow=c(5, 3)) # Set up a 4x3 grid for plotting
for (i in 1:15) {
  plot(ts_airline[, i], main = colnames(ts_airline)[i], xlab = "Year", ylab = colnames(ts_airline)[i], type = "l")
}
```

#### Narrowing Dataset

We are choosing to focus on the domestic changes in air traffic. For this, we will explore the Domestic Passengers, Domestic Flight Revenue, and the Domestic Load Factor.

```{r}
plot(ts_airline[, "Dom_Pax"], main = "Domestic Passenger", xlab = "Year", ylab = "Passenger Count", type = "l")

plot(ts_airline[, "Dom_RPM"], main = "Domestic Flight Revenue", xlab = "Year", ylab = "Revenue", type = "l")

plot(ts_airline[, "Dom_LF"], main = "Domestic Load Factors", xlab = "Year", ylab = "Load Factor", type = "l")
```
##### Visualizing Subset of Data Pertaining to COVID-19
We chose to take a closer look at the time period of Jan 2020 to May 2023, the dates of the Covid-19 pandemic.
```{r}
ts_subset <- window(ts_airline, start = c(2020, 1), end = c(2023, 5))

plot(ts_subset[, "Dom_Pax"], main = "Domestic Passenger Count (Jan 2020 - May 2023)", xlab = "Year", ylab = "Passenger Count", type = "l")

plot(ts_subset[, "Dom_RPM"], main = "Domestic Flight Revenue (Jan 2020 - May 2023)", xlab = "Year", ylab = "Revenue", type = "l")

plot(ts_subset[, "Dom_LF"], main = "Domestic Load Factors (Jan 2020 - May 2023)", xlab = "Year", ylab = "Load Factor", type = "l")
```

#### Decomposition
```{r}
decomp_dom_pax <- decompose(ts_airline[, "Dom_Pax"])

plot(decomp_dom_pax)

decomp_dom_rpm <- decompose(ts_airline[, "Dom_RPM"])

plot(decomp_dom_rpm)

decomp_dom_lf <- decompose(ts_airline[, "Dom_LF"])

plot(decomp_dom_lf)
```
#### Prepping Data for Model Fitting

```{r}
pre_pandemic_start <- c(2003, 1)  # January 2003
pre_pandemic_end <- c(2019, 12)   # December 2019
pandemic_start <- c(2020, 1)      # January 2020
pandemic_end <- c(2023, 5)        # May 2023
post_pandemic_end <- c(2023, 9)   # September 2023 (End of dataset)

pre_pandemic_data = window(ts_airline, start = pre_pandemic_start, end = pre_pandemic_end)
pandemic_data = window(ts_airline, start = pandemic_start, end = pandemic_end)
post_pandemic_data = window(ts_airline, start = pandemic_end, end = post_pandemic_end)

dim(pre_pandemic_data)
dim(pandemic_data)
dim(post_pandemic_data)
```
Splitting the pre-pandemic and pandemic data into training and testing chunks and then aggregating to generate a consolidated training and testing dataset.

```{r}
num_obs_pre_train = 163
num_obs_pre_test = 41
num_obs_pandemic_train = 32
num_obs_pandemic_test = 9
num_obs_post_test = 5

pre_train_data <- window(pre_pandemic_data, start = c(2003, 1), end = c(2016, 7))
pre_test_data <- window(pre_pandemic_data, start = c(2016, 8), end = c(2019, 12))
pandemic_train_data <- window(pandemic_data, start = c(2020, 1), end = c(2022, 8))
pandemic_test_data <- window(pandemic_data, start = c(2022, 9), end = c(2023,5))
post_test_data = post_pandemic_data

dim(pre_train_data)
dim(pre_test_data)
dim(pandemic_train_data)
dim(pandemic_test_data)
dim(post_test_data)

train_data = rbind(pre_train_data, pandemic_train_data)
test_data = rbind(pre_test_data, pandemic_test_data, post_test_data)

dim(train_data)
dim(test_data)
```
By doing this, we have prepped our data set to be used for training various models and testing the accuracy while taking into account the large disturbance in data that was caused due to COVID-19

#### Analyzing Stationarity of Data
##### Domestic Passengers
```{r}
ts_pre_pandemic <- window(ts_airline, start = c(2003,1), end = c(2019,12))

ur.kpss(ts_airline[, "Dom_Pax"]) %>% summary()      #Stationary
ur.kpss(ts_pre_pandemic[, "Dom_Pax"]) %>% summary() #Non-Stationary
```
Based on the KPSS test, we identified that the entire dataset is stationary, but the subset of pre-pandemic data is not stationary.

```{r}
autoplot(acf(ts_airline[, "Dom_Pax"]))
autoplot(acf(ts_pre_pandemic[, "Dom_Pax"]))

```
These ACF graphs reconfirm the results of the KPSS test.
```{r}

plot(density(ts_airline[, "Dom_Pax"]), main = "Density Plot of Dom_Pax Time Series (2003 - 2023)", xlab = "Dom_Pax Values", ylab = "Density")
plot(density(ts_pre_pandemic[, "Dom_Pax"]), main = "Density Plot of Dom_Pax Time Series (ts_pre_pandemic)", xlab = "Dom_Pax Values", ylab = "Density")

```
These density plots provide a visual look into how the data is distributed allowing us to better understand the nature of the data.
```{r}
ggsubseriesplot(ts_airline[, "Dom_Pax"])
ggsubseriesplot(ts_pre_pandemic[, "Dom_Pax"])

```
These graphs reiterate the seasonality of the data.


##### Domestic Revenue Passenger Miles
```{r}

ur.kpss(ts_airline[, "Dom_RPM"]) %>% summary()      #Non-Stationary (5% Level)
ur.kpss(ts_pre_pandemic[, "Dom_RPM"]) %>% summary() #Non-Stationary

```
Based on the KPSS test, we identified that both the entire dataset and the subset of pre-pandemic data are not stationary given a 5% significance level.

```{r}
autoplot(acf(ts_airline[, "Dom_RPM"]))
autoplot(acf(ts_pre_pandemic[, "Dom_RPM"]))

```
These ACF graphs reconfirm the results of the KPSS test.

```{r}
plot(density(ts_airline[, "Dom_RPM"]), main = "Density Plot of Dom_RPM Time Series (2003 - 2023)", xlab = "Dom_RPM Values", ylab = "Density")
plot(density(ts_pre_pandemic[, "Dom_RPM"]), main = "Density Plot of Dom_RPM Time Series (ts_pre_pandemic)", xlab = "Dom_RPM Values", ylab = "Density")

```
These density plots provide a visual look into how the data is distributed allowing us to better understand the nature of the data.

```{r}
ggsubseriesplot(ts_airline[, "Dom_RPM"])
ggsubseriesplot(ts_pre_pandemic[, "Dom_RPM"])

```
These graphs reiterate the seasonality of the data.

##### Domestic Load Factor
```{r}
ur.kpss(ts_airline[, "Dom_LF"]) %>% summary()       #Stationary
ur.kpss(ts_pre_pandemic[, "Dom_LF"]) %>% summary()  #Non-Stationary
```
Based on the KPSS test, we identified that the entire dataset is stationary, but the subset of pre-pandemic data is not stationary.

```{r}
autoplot(acf(ts_airline[, "Dom_LF"]))
autoplot(acf(ts_pre_pandemic[, "Dom_LF"]))

```
These ACF graphs reconfirm the results of the KPSS test.

```{r}
plot(density(ts_airline[, "Dom_LF"]), main = "Density Plot of Dom_LF Time Series (2003 - 2023)", xlab = "Dom_LF Values", ylab = "Density")
plot(density(ts_pre_pandemic[, "Dom_LF"]), main = "Density Plot of Dom_LF Time Series (Pre-Pandemic)", xlab = "Dom_LF Values", ylab = "Density")
```
These density plots provide a visual look into how the data is distributed allowing us to better understand the nature of the data.

```{r}
ggsubseriesplot(ts_airline[, "Dom_LF"])
ggsubseriesplot(ts_pre_pandemic[, "Dom_LF"])

```
These graphs reiterate the seasonality of the data.

## Future Plans

- We believe that because of the non-stationarity of the data, we will have to use differencing to make it stationary. 
  - This will be useful when we try to answer the research question relation to correlation
- We will then proceed to use a variety of models to predict/forecast different variables
  - We will use models like: Linear Regression, Quadratic Regression, Logistic Regression, ARIMA, SARIMA etc.
  - We will see how those models compare to naive models like SES, MES, and Holt/Holt-Winters.
- We will then assess how accurate we are and discuss how the COVID-19 pandemic affected our modeling.
  - We will use various accuracy metrics including: MAPE, MASE, Residual analysis

