---
title: "Airline_Traffic_TS"
output: html_document
date: "2024-04-29"
---

This is part 2 of the Airline Traffic Time Series Project.

By: Harshith Samayamantula, Devan Pandya
4/29/24
Time Series 

```{r}
require(zoo)
require(ggplot2)
require(forecast)
require(TTR)
library(fpp2)
require(lubridate)
library(tseries)
library(urca)

airline_dataset = read.csv("air traffic.csv")
```

## Research Question
Some research questions we want to explore include the following:
- Can we accurately forecast future air travel demand based on historical data? Specifically, passenger numbers and revenue passenger-miles?
  - The result is testing various methods for forecasting accuracy and discussing the level of accuracy we're able to attain.
- Is there a correlation between the number of flights (Flt) and total passenger volume (Pax)? How do revenue passenger-miles (RPM) correlate with the number of flights and passenger volume?
  - Correlation will be shown through autocorrelation and partial autocorrelation analysis, identifying if the data is stationary, and running other tests.
  
```{r}
print(airline_dataset)
```


```{r}
airline_dataset$Year_Month = as.yearmon(paste(airline_dataset$Year, airline_dataset$Month), "%Y %m")
```

```{r}
print(airline_dataset)
str(airline_dataset)
```
```{r}
# Convert character columns to numeric columns
airline_dataset$Dom_Pax <- as.numeric(gsub(",", "", airline_dataset$Dom_Pax))
airline_dataset$Int_Pax <- as.numeric(gsub(",", "", airline_dataset$Int_Pax))
airline_dataset$Pax <- as.numeric(gsub(",", "", airline_dataset$Pax))
airline_dataset$Dom_Flt <- as.numeric(gsub(",", "", airline_dataset$Dom_Flt))
airline_dataset$Int_Flt <- as.numeric(gsub(",", "", airline_dataset$Int_Flt))
airline_dataset$Flt <- as.numeric(gsub(",", "", airline_dataset$Flt))
airline_dataset$Dom_RPM <- as.numeric(gsub(",", "", airline_dataset$Dom_RPM))
airline_dataset$Int_RPM <- as.numeric(gsub(",", "", airline_dataset$Int_RPM))
airline_dataset$RPM <- as.numeric(gsub(",", "", airline_dataset$RPM))
airline_dataset$Dom_ASM <- as.numeric(gsub(",", "", airline_dataset$Dom_ASM))
airline_dataset$Int_ASM <- as.numeric(gsub(",", "", airline_dataset$Int_ASM))
airline_dataset$ASM <- as.numeric(gsub(",", "", airline_dataset$ASM))
str(airline_dataset)
```

#### Converting to TS
```{r}
ts_airline <- ts(airline_dataset[, c("Dom_Pax", "Int_Pax", "Pax", "Dom_Flt", "Int_Flt", "Flt", "Dom_RPM", "Int_RPM", "RPM", "Dom_ASM", "Int_ASM", "ASM", "Dom_LF", "Int_LF", "LF")], start = c(2003, 1), frequency = 12)
```


```{r}
print(window(ts_airline, start = c(2003, 1), end = c(2003, 5)))
```

#### Visualizing Data
```{r}
par(mar = c(3, 3, 1, 1))
par(mfrow=c(5, 3)) # Set up a 4x3 grid for plotting
for (i in 1:15) {
  plot(ts_airline[, i], main = colnames(ts_airline)[i], xlab = "Year", ylab = colnames(ts_airline)[i], type = "l")
}
```

#### Narrowing Dataset

We are choosing to focus on the domestic changes in air traffic. For this, we will explore the Domestic Passengers, Domestic Flight Revenue, and the Domestic Load Factor.

```{r}
plot(ts_airline[, "Dom_Pax"], main = "Domestic Passenger", xlab = "Year", ylab = "Passenger Count", type = "l")

plot(ts_airline[, "Dom_RPM"], main = "Domestic Flight Revenue", xlab = "Year", ylab = "Revenue", type = "l")

plot(ts_airline[, "Dom_LF"], main = "Domestic Load Factors", xlab = "Year", ylab = "Load Factor", type = "l")
```

##### Visualizing Subset of Data Pertaining to COVID-19
We chose to take a closer look at the time period of Jan 2020 to May 2023, the dates of the Covid-19 pandemic.
```{r}
ts_subset <- window(ts_airline, start = c(2020, 1), end = c(2023, 5))

plot(ts_subset[, "Dom_Pax"], main = "Domestic Passenger Count (Jan 2020 - May 2023)", xlab = "Year", ylab = "Passenger Count", type = "l")

plot(ts_subset[, "Dom_RPM"], main = "Domestic Flight Revenue (Jan 2020 - May 2023)", xlab = "Year", ylab = "Revenue", type = "l")

plot(ts_subset[, "Dom_LF"], main = "Domestic Load Factors (Jan 2020 - May 2023)", xlab = "Year", ylab = "Load Factor", type = "l")
```

###### Analysis of unsual observations
Prior to the pandemic, generally domestic passenger count, domestic flight revenue, and domestic load factors had steadily increased over time. Overall, again excluding the pandemic era, all three data sets have strong seasonality, with the summer months being the busiest time for domestic air travel. Domestic passenger count and flight revenue experience a slight decrease in 2008, most likely due to the Great Recession. However, the airline industry began to bounce back by 2010. The Covid-19 pandemic represents the greatest anomaly in the data set as passenger traffic and revenues fell to record lows during the early months of the pandemic. Wit the widespread distribution of vaccines and subsequent reductions on travel restrictions, domestic air travel began to bounce back in early 2021. However, it wasn't until 2023 when domestic travel data return to their pre-pandemic levels. As the world continues to move past to the pandemic, it will be interesting to see how domestic air travel behaves as the U.S. grapples with rising fuel costs, a shortage of airplanes and staff, and soaring passenger demand. 